---
title: "R Project #1"
author: "David T Boyd"
format: html
editor: visual
---

## Exploratory Data Analysis (EDA): NYC 311 Service Requests

### **Objective**

Self-contained report that ingests the raw CSV, profiles missing data, cleans column types, and produces a narrative EDA with `ggplot2` — complaint volume by borough, top complaint types, and seasonal trends.

### Thought Process

Load necessary packages

```{r}
library(httr)
library(jsonlite)
library(socratadata)
library(data.table)
library(dplyr)
library(bit64)
#library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
library(tigris)
library(janitor)

```

Load the data, perform a summary analysis to understand data, structure and outliers

```{r}

df_311 <- fread("/home/davidtboyd/Data-Analysis-Portfolio-R/ExploratoryDataAnalysis/311_Service_Requests_from_2020_to_Present_20260509.csv")



```

Clean column names to standard names with underscores instead of spaces

```{r}
df_311 <- clean_names(df_311)
```

Get df header information (20 rows) and structure.

```{r}
head(df_311, 20)
str(df_311)

```

Check non-character columns and determine if any can / should be converted to chr. In this example, Unique Key and BBL can be switched to chr. There are no calculations that would be performed on these

```{r}

Filter(Negate(is.character), df_311)  # returns the actual columns, not just their types
```

Convert ID fields into 'chr' data types and convert date fields to date types

```{r}
df_311 <- df_311 |> mutate(
  `unique_key` = as.character(`unique_key`),
  bbl = as.character(bbl),
  'created_date' = mdy_hms(`created_date`),
  'closed_date' = mdy_hms(`closed_date`),
  'due_date' = mdy_hms(`due_date`),
  'resolution_action_updated_date' = mdy_hms(`resolution_action_updated_date`)
    )

```

From the output of str(df_311) need to look at columns that appear to have excessive null values e.g. Vehicle Type, Taxi Company Borough, etc.. Based on the output only BBL needs to be addressed.

```{r}
na_pct <- colMeans(is.na(df_311)) * 100
na_pct[na_pct > 2]

```

Let's find the minimum and maximum date, based on Created Date

```{r}
start_date = min(df_311$`created_date`, na.rm = TRUE)
finish_date = max(df_311$`created_date`, na.rm = TRUE)
print(start_date)
print(finish_date)

```

Let's export 1000 rows to a csv file for inspection in Excel if necessary

```{r}
set.seed(76)
top1000 <- df_311 %>% slice_sample(n = 1000)
write.csv(top1000, "311_Sample.csv", row.names = FALSE)

```

1st Request: Complaint Volume by Borough

We need to create a new df for complaints filtered by agency, so I will pull the distinct values for Agency

```{r}
df_agencies <- df_311 %>% distinct(`agency`)
df_boroughs <- df_311 |> distinct(`borough`)
print(df_agencies)
print(df_boroughs)


```

based on this list, we will be seeking only calls for NYPD.

```{r}
df_complaints <- filter(df_311, agency == "NYPD")

```

Created basic bar chart

```{r}
df_311 %>%
  group_by(borough) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = borough, y = count)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_number(scale = 1/1000000, suffix = "M", accuracy = 0.1))
```

```{r}
borough_counts <- df_311 %>%
  filter(!is.na(borough), borough != "Unspecified") %>%
  mutate(borough = tolower(borough)) %>%
  count(borough, name = "calls")

nyc_boroughs <- counties(state = "NY", cb = TRUE, class = "sf") %>%
  filter(NAME %in% c("New York", "Bronx", "Queens", "Kings", "Richmond")) %>%
  mutate(Borough = case_when(
    NAME == "New York" ~ "MANHATTAN",
    NAME == "Kings"    ~ "BROOKLYN",
    NAME == "Richmond" ~ "STATEN ISLAND",
    TRUE               ~ toupper(NAME)
  ))

clean_names(nyc_boroughs)

nyc_choropleth <- nyc_boroughs %>%
  mutate(Borough = tolower(Borough)) %>%
  left_join(borough_counts, by = c("Borough" = "borough"))

ggplot(nyc_choropleth) +
  geom_sf(aes(fill = calls), color = "white", linewidth = 0.6) +
  geom_sf_label(aes(label = paste0(Borough, "\n", comma(calls))),
                size = 3, label.size = 0) +
  scale_fill_viridis_c(name = "311 Calls", labels = comma) +
  theme_void() +
  labs(title = "NYC 311 Calls by Borough")
```

The previous map uses black text on all shadings which makes Stanten Island difficult to read. Updated code to use white lettering with darker backgrounds

```{r}
nyc_choropleth <- nyc_choropleth %>%
  mutate(
    scaled = (calls - min(calls, na.rm = TRUE)) / 
             (max(calls, na.rm = TRUE) - min(calls, na.rm = TRUE)),
    text_color = if_else(scaled < 0.4, "white", "black")
  )

ggplot(nyc_choropleth) +
  geom_sf(aes(fill = calls), color = "white", linewidth = 0.6) +
  geom_sf_label(
    aes(label = paste0(Borough, "\n", comma(calls)), color = text_color),
    size = 3, label.size = 0
  ) +
  scale_fill_viridis_c(name = "311 Calls", labels = comma) +
  scale_color_identity() +  # tells ggplot to use the color values as-is
  theme_void() +
  labs(title = "NYC 311 Calls by Borough")
```

Test the Seasonal-Trends branch
